An open and continuously updated fern tree of life (FTOL)

### 2022-02-14 ms/manuscript.Rmd
Rearrange functions, calculate ages in targets plan

# Load functions
source(here::here("R/ms_functions.R"))
source(here::here("R/functions.R"))

# Specify percentage format
percent <- function(...) {scales::percent(...)}
number <- function(...) {scales::number(big.mark = ",", ...)}
```

@@

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files,
		other_dates
	),
	store = here::here("_targets")
)
```

@@

# Make tibble with one row per species in Sanger sampling
sanger_sampling <- make_sanger_sampling_tbl(
	plastome_metadata_renamed,
	sanger_tree = plastid_tree_dated,
	ppgi_taxonomy = ppgi_taxonomy)

@@

# Helper function to add major_clade:
# a "rank" that may be at order or suborder level for plotting
add_major_clade <- function(data) {
	data %>%
		mutate(
			major_clade = coalesce(suborder, order),
			major_clade = case_when(
				major_clade %in% c("Ophioglossales", "Psilotales") ~ 
					"Ophioglossales + Psilotales",
				TRUE ~ major_clade
			)
		)
}

# Tally FTOL sampling by rank
ftol_sampling_by_rank <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	# add major_clade
	add_major_clade() %>%
	# drop suborder
	select(-suborder) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup

@@

```{r topology-stats}
#' Function to get support value in BS%
#'
#' @param phy an object of class "phylo".
#' @param tips a vector of mode numeric or character specifying the tips
#'
#' @return Bootstrap support as % of that node
get_bs <- function(phy, tips) {
	node_select <- ape::getMRCA(phy, tips)
	as_tibble(phy) %>%
		filter(node == node_select) %>%
		pull(label) %>%
		parse_number() %>%
		magrittr::multiply_by(0.01) %>%
		percent()
}

ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
hym_gle_bs <- get_bs(plastome_tree, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
cyath_salv_bs <- get_bs(plastome_tree, c("Plagiogyria_adnata", "Azolla_filiculoides"))
denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))
```

@@

```{r divergence-times-stats}
# Get overall tree height
tree_height <- max(phytools::nodeHeights(plastid_tree_dated))

# Define function for calculating age based on two genera
get_age_from_genus_pair <- function(tip_pair) {
	sanger_sampling %>%
	filter(genus %in% tip_pair) %>%
	group_by(genus) %>%
	slice(1) %>%
	pull(species) %>%
  ape::getMRCA(plastid_tree_dated, .) %>%
	phytools::nodeheight(plastid_tree_dated, .) %>%
	magrittr::subtract(tree_height, .) %>%
	number(accuracy = 0.1)
}

# Calculate age of various clades
fern_age <- get_age_from_genus_pair(c("Equisetum", "Polypodium"))
lepto_age <- get_age_from_genus_pair(c("Osmunda", "Polypodium"))
polypodiales_age <- get_age_from_genus_pair(c("Lindsaea", "Polypodium"))
eupoly_i_age <- get_age_from_genus_pair(c("Hypodematium", "Polypodium"))
eupoly_ii_age <- get_age_from_genus_pair(c("Cystopteris", "Asplenium"))

# Get ages for families
sanger_sampling_family <- select(sanger_sampling, species, family)

# Only use monophyletic families
mono_test_family <- assess_monophy(
	taxon_sampling = sanger_sampling_family,
	tree = plastid_tree_dated,
	tax_levels = "family"
) %>%
	get_result_monophy(1)

family_ages <-
mono_test_family %>%
	filter(monophyly == "Yes") %>%
	mutate(
		mrca = parse_number(mrca),
		stem_node = map_dbl(
			mrca,
			~phangorn::Ancestors(plastid_tree_dated, ., type = "parent"))
	) %>%
	select(family = taxon, stem_node) %>%
mutate(
	# Height is distance above root
	height = map_dbl(stem_node, ~phytools::nodeheight(
		tree = plastid_tree_dated, node = .))
) %>%
	mutate(
		# Age is the total length of the tree minus height
		age = tree_height - height
	)
```

We estimate the crown age of ferns to be `r fern_age` ma.
Ages of other major crown groups are leptosporangiates (`r lepto_age` ma), Polypodiales (`r polypodiales_age` ma), eupolypods I (`r eupoly_i_age` ma), and eupolypods II (`r eupoly_ii_age` ma).
Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
Estimated stem ages of fern families were generally older than previous studies. (`r figure("divtimes")`).

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
family_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -family) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median"))

# Convert family errors to almost long format
family_ages_errors <-
	family_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	select(family, matches("ts|rothfels")) %>%
	select(-matches("median"))

@@

div_times_plot <-
ggplot() +
	geom_errorbarh(
		data = family_ages_errors,
		aes(xmax = high, xmin = low, color = author, y = family),
		position = position_dodge(width = 0.5)
	) +
	geom_point(
		data = filter(family_ages_median, age_type %in% c("best", "median")),
		aes(x = age, y = family, color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = ages_shapes,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_color_manual(
		values = ages_colors,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	labs(x = "Age (millions of years)") +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank(),
		legend.position = "bottom"
	)
```

>>>>

# Load functions
source(here::here("R/ms_functions.R"))
source(here::here("R/functions.R"))
```

@@

# Targets from MS workflow
tar_load(
	c(
		sanger_sampling,
		gb_species_by_year,
		ref_files,
		other_dates,
		crown_ages,
		family_stem_ages
	),
	store = here::here("_targets")
)
```

@@

@@

# Tally FTOL sampling by rank
ftol_sampling_by_rank <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	# drop suborder
	select(-suborder) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup

@@

```{r topology-stats}
ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
hym_gle_bs <- get_bs(plastome_tree, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
cyath_salv_bs <- get_bs(plastome_tree, c("Plagiogyria_adnata", "Azolla_filiculoides"))
denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))
```

@@

We estimate the crown age of ferns to be `r crown_ages$ferns` ma.
Ages of other major crown groups are leptosporangiates (`r crown_ages$leptos` ma), Polypodiales (`r crown_ages$polypodiales` ma), eupolypods I (`r crown_ages$eupoly_i` ma), and eupolypods II (`r crown_ages$eupoly_ii` ma).
Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
Estimated stem ages of fern families were generally older than previous studies (`r figure("divtimes")`).

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
family_stem_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -family) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median"))

# Convert family errors to almost long format
family_ages_errors <-
	family_stem_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	select(family, matches("ts|rothfels")) %>%
	select(-matches("median"))

@@

div_times_plot <-
ggplot(mapping = aes(y = family)) +
	geom_errorbarh(
		data = family_ages_errors,
		aes(xmax = high, xmin = low, color = author),
		position = position_dodge(width = 0.5)
	) +
	geom_point(
		data = filter(family_ages_median, age_type %in% c("best", "median")),
		aes(x = age, color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = ages_shapes,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_color_manual(
		values = ages_colors,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	labs(x = "Age (millions of years)") +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank(),
		legend.position = "bottom"
	)
```
